The Weakly Nonlinear Magnetorotational Instability in a local geometry

### 2017-03-22 paper/thingap_ref_response/thingap_mri_paper_ref_response_apj.tex
minor edits and some additiona text.

Because we nondimensionalize $B$ by the magnitude of the background field strength, $B_0 \equiv 1$ in Equations \ref{eqset1} - \ref{eqset4}. To excite the weakly nonlinear MRI, we tune the background magnetic field away from stability. %(recall that stronger vertical fields stabilize the MRI). 
We do so by substituting \edit1{$B = B_0\left(1 + \epsilon^2\right)$}. The degree of departure from the marginal state is measured by the small parameter $\epsilon$. An $\mathcal{O}\left(\epsilon^2\right)$ \edit1{strengthening} of the background magnetic field destabilizes a finite band of wave modes with a width of $\mathcal{O}\left(\epsilon\right)$, which interact nonlinearly. \edit1{We note that this definition of $\epsilon$ is opposite in sign to nearly all previous works \citep[e.g.][]{Umurhan:2007dz,Umurhan:2007hs}. Because in the ideal limit, the MRI can be tuned into instability by setting $B_0$ to its critical value and then \emph{decreasing} its value, it is natural to consider $\epsilon^2$ as a weakening of the background field. However, as we show in figure~\ref{fig:growth_rate_vs_B0}, for the dissipative case with $\eta, \nu \ne 0$, when all other parameters are critical, decreasing $B_0$ leads to stability, while increasing it pushes the system into instability. Figure~\ref{fig:growth_rate_vs_B0} is symmetric about $B_0 = 0$, as it must be, since the MRI is insensitive to the sign of the background field. There are several places at which the derivative of $\gamma$ appears discontinuous; this is not physical but rather reflects the fact that we define $\gamma$ as the growth rate of the most unstable mode. That is, it is the maximum real part of the eigenvalues of the linearized system (e.g. equation~\ref{eq:unperturbed_matrix_equations} with $\mathbf{N} = 0$). Because there are four wave families in rotating incompressible MHD, each modified differently by changing $B_0$, when the growth rates of the individual modes cross, there appear piecewise continuous solutions. We highlight one such point in the inset in figure~\ref{fig:growth_rate_vs_B0}, where the MRI mode becomes more stable than another mode which is always stable. Since all of these piecewise discontinuities are below $\gamma=0$, they do not affect the analysis here.}

@@

\edit1{Many studies of the MRI consider the instability in a shearing box, i.e. a wall-less local approximation that is meant to represent a small section of a disk. The shearing box obtains when Equations \ref{eqset1} - \ref{eqset4} are subjected to shear periodic radial boundary conditions rather than Equation \ref{eq:wallbcs}. The periodic nature of the shearing box allows us to decompose the fluid perturbations into Fourier modes proportional to $e^{i k_x x + i k_z z}$. This makes the shearing box MRI straightforward to treat analytically. However, as noted above, the fastest-growing linear MRI modes in the shearing box are also exact solutions of the nonlinear MRI equations -- that is, $J(\hat{\psi_0}, \hat{\psi_0}) = J(\hat{\psi_0}, \nabla^2 \hat{\psi_0}) = 0$ for $\hat{\psi_0} \propto e^{i k_x x + i k_z z}$. While this may be an appealing trait for analytic simplicity, it leads to the unphysical conclusion that the fastest growing modes will never nonlinearly interact \citep{Goodman:1994ul}. This `nonlinear property' will not be satisfied for two MRI modes with nonparallel wavenumbers, but with vertically periodic boundary conditions and a vertical background magnetic field the most unstable mode has a strictly axial vertical wavenumber. Thus a formal weakly nonlinear analysis cannot be conducted, as the most unstable mode will never nonlinearly interact with itself or its complex conjugate. Similarly, we cannot analytically examine interactions between MRI channel modes and damped eigenmodes belonging to other wave families. This is analytically tracked for other plasma instabilities by tracking the amplitudes of growing, marginal, and damped eigenmodes simultaneously \citep[e.g.][]{Makwana:2011}. While the shearing box approximation allows the projection of the perturbed MRI equations into the basis set of linear eigenmodes, nonlinear coupling between modes will remain zero.}

@@

\edit1{We analyze the average energy and angular momentum transport in the simulation domain (Figure \ref{fig:sim_theory_energy_jdot}).  }

@@

\section{Acknowledgments}
S.E.C. was supported by a National Science Foundation Graduate Research Fellowship under grant No. DGE-16-44869. J.S.O. acknowledges support from NASA grant \edit1{NNX16AC92G}. \edit1{We thank the anonymous referee for thoughtful comments that greatly improved the manuscript.} The authors thank Mordecai Mac Low, Jeremy Goodman, John Krommes, Geoff Vasil, and Ellen Zweibel for useful discussion. \software{Dedalus: http://dedalus-project.org/}

@@

>>>>

Because we nondimensionalize $B$ by the magnitude of the background field strength, $B_0 \equiv 1$ in Equations \ref{eqset1} - \ref{eqset4}. To excite the weakly nonlinear MRI, we tune the background magnetic field away from stability. %(recall that stronger vertical fields stabilize the MRI). 
We do so by substituting \edit1{$B = B_0\left(1 + \epsilon^2\right)$}. The degree of departure from the marginal state is measured by the small parameter $\epsilon$. An $\mathcal{O}\left(\epsilon^2\right)$ \edit1{strengthening} of the background magnetic field destabilizes a finite band of wave modes with a width of $\mathcal{O}\left(\epsilon\right)$, which interact nonlinearly. \edit1{We note that this definition of $\epsilon$ is opposite in sign to nearly all previous works \citep[e.g.][]{Umurhan:2007dz,Umurhan:2007hs}. Because in the ideal limit, the MRI can be tuned into instability by setting $B_0$ to its critical value and then \emph{decreasing} its value, it is natural to consider $\epsilon^2$ as a weakening of the background field \citep[as is done correctly in][for example]{Vasil:2015}. However, as we show in figure~\ref{fig:growth_rate_vs_B0}, for the \emph{dissipative} case with $\eta, \nu \ne 0$, when all other parameters are critical, decreasing $B_0$ leads to stability, while increasing it pushes the system into instability. Figure~\ref{fig:growth_rate_vs_B0} is symmetric about $B_0 = 0$, as it must be, since the MRI is insensitive to the sign of the background field. There are several places at which the derivative of $\gamma$ appears discontinuous; this is not physical but rather reflects the fact that we define $\gamma$ as the growth rate of the most unstable mode. That is, it is the maximum real part of the eigenvalues of the linearized system (e.g. equation~\ref{eq:unperturbed_matrix_equations} with $\mathbf{N} = 0$). Because there are four wave families in rotating incompressible MHD, each modified differently by changing $B_0$, when the growth rates of the individual modes cross, there appear piecewise continuous solutions. We highlight one such point in the inset in figure~\ref{fig:growth_rate_vs_B0}, where the MRI mode becomes more stable than another mode which is always stable. Since all of these piecewise discontinuities are below $\gamma=0$, they do not affect the analysis here.}

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{thingap_stressfig_Pm_1E-2}
\caption{\edit1{Total (Reynolds + Maxwell) stress in the domain as predicted from the weakly nonlinear theory.}} \label{fig:wnl_stress}
\end{figure}

\edit1{Many studies of the MRI consider the instability in a shearing box, i.e. a wall-less local approximation that is meant to represent a small section of a disk. The shearing box is the limit in which Equations \ref{eqset1} - \ref{eqset4} are subjected to shear periodic radial boundary conditions rather than Equation \ref{eq:wallbcs} \citep[e.g.][]{2008A&A...481...21R}. The periodic nature of the shearing box allows us to decompose the fluid perturbations into Fourier modes proportional to $e^{i k_x x + i k_z z}$. This makes the shearing box MRI straightforward to treat analytically. However, as noted above, the fastest-growing linear MRI modes in the shearing box are also exact solutions of the nonlinear MRI equations -- that is, $J(\hat{\psi_0}, \hat{\psi_0}) = J(\hat{\psi_0}, \nabla^2 \hat{\psi_0}) = 0$ for $\hat{\psi_0} \propto e^{i k_x x + i k_z z}$. While this may be an appealing trait for analytic simplicity, it leads to the unphysical conclusion that the fastest growing modes will never nonlinearly interact \citep{Goodman:1994ul}. This `nonlinear property' will not be satisfied for two MRI modes with nonparallel wavenumbers, but with vertically periodic boundary conditions and a vertical background magnetic field the most unstable mode has a strictly axial vertical wavenumber. Thus a formal weakly nonlinear analysis cannot be conducted, as the most unstable mode will never nonlinearly interact with itself or its complex conjugate. Similarly, we cannot analytically examine interactions between MRI channel modes and damped eigenmodes belonging to other wave families. This is analytically tracked for other plasma instabilities by tracking the amplitudes of growing, marginal, and damped eigenmodes simultaneously \citep[e.g.][]{Makwana:2011}. While the shearing box approximation allows the projection of the perturbed MRI equations into the basis set of linear eigenmodes, nonlinear coupling between modes will remain zero.}

@@

\edit1{We analyze the average energy and angular momentum transport in the simulation domain (Figure \ref{fig:sim_theory_energy_jdot}). The theory slightly overpredicts both quantities, though with a reasonably good agreement considering that the weakly nonlinear theory applies rigorously to a channel of infinite height, while the simulation was carried out in a box with a vertical extent of only two critical wavelengths. We defer further comparisons between simulation and theory, including an analysis of the effect of the box height on the simulated flow, to future work. }

@@

\section{Acknowledgments}
S.E.C. was supported by a National Science Foundation Graduate Research Fellowship under grant No. DGE-16-44869. J.S.O. acknowledges support from NASA grant \edit1{NNX16AC92G}. \edit1{We thank the anonymous referee for thoughtful comments that greatly improved the manuscript.} \edit1{We also} thank Mordecai Mac Low, Jeremy Goodman, John Krommes, Geoff Vasil, and Ellen Zweibel for useful discussion. \software{Dedalus: http://dedalus-project.org/}

@@

\bibitem[Regev \& Umurhan(2008)]{2008A&A...481...21R} Regev, O., \& Umurhan, O.~M.\ 2008, \aap, 481, 21
